---
title: 'ML Final Project: Analysis'
author: "Andrew Pagtakhan"
date: "January 20, 2020"
output: pdf_document
---

```{r setup, include=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(dendextend)
library(GGally)
library(ggfortify)
library(ggplot2)
library(ggrepel)
library(gridExtra)
library(knitr)
library(mclust)
library(NbClust)

# set for reproducible results
set.seed(14)

# clear variables
rm(list = ls())

# set working directory
# wd <- 'C:/Users/apagta950/Documents/NYU/Courses/Spring 2020/ML/Projects/ML-NBA/Data/'
opts_knit$set(root.dir = normalizePath('..'))

# define file name for analysis
filename <- 'Data/season_stats_clean.csv'
```

# Research Question: Are there underlying patterns of groupings between NBA team compensation vs. overall team skillsets?

## Data Cleaning
The data cleaning was done in separate R scripts. 
https://github.com/joemarlo/ML-NBA
Steps:
* Filtered data to the 2016 - 2017 season
* Assigned player to one team based on the most minutes he played for, including stats across all teams played for
* Scraped player salaries and RPM data from ESPN website, using fuzzy matching on player names to join to the main dataset
* Scaled/Transformed data using cube root (shown below)

## Exploratory Data analysis
```{r}
# load data
nba <- read.csv(filename)
str(nba)
```

```{r}
# replace NA values in RPM column with 0s
nba$RPM[is.na(nba$RPM)] <- 0
```

We decided to use the 14 features in our analysis (on a per game basis) because they provide a good balance of offensive (e.g Pts, Ast) and defensive stats (e.g. Reb, Blk). Additionally, advanced statistics such as VORP and PER utilize a combination of these 'base' statistics in their calculations. So, by using these base statistics between offensive and defensive metrics, this is more likely to provide more balanced groupings between those who are more offensive and those who are better at defense.
```{r}
# plot distributions of key per game stats
features_pg <- grep('_pg', names(nba), value = TRUE)
nba_feat <- nba[ , features_pg]

# further subset for plotting purposes
feat_plot <- c('MP_pg', 'PTS_pg', 'AST_pg', 'TRB_pg', 
               'STL_pg', 'BLK_pg')

# dataframe for plotting
nba_feat_plot <- nba[ , feat_plot]

# plot key stats
plots_stats <- list(rep(NA, length = length(feat_plot)))
for (i in seq_along(feat_plot)) {
  p <- ggplot(nba_feat_plot, aes_string(x = feat_plot[i])) + 
        geom_density(fill = 'lightblue', alpha = 0.4) + 
        ggtitle(feat_plot[i]) + 
        # center title
        theme(plot.title = element_text(hjust = 0.5))
  plots_stats[[i]] <- p
}

# multiple plotting
cowplot::plot_grid(plotlist = plots_stats, nrow = 2) 
```


```{r, warning=FALSE}
# look at pairs plots for key stats
nba_feat_plot_pos <- cbind(nba_feat_plot, nba$Pos)
ggpairs(nba_feat_plot_pos,
        columns = 1:(length(names(nba_feat_plot_pos)) - 1),
        progress = FALSE, 
        mapping = ggplot2::aes(colour = nba$Pos, alpha = 0.4),
        upper = list(continuous = wrap('cor', size = 0))
)
```

Most of the features such as points and assists look like they exhibit a negative binomial distribution. If we look at the pairs plot, it looks like there are clear groupings by position.

## Run PCA to inspect if there are any groupings

Before running any clustering algorithms, we will perform Principal Component Analysis to determine if there are any potential clusters. We first scaled the data because the ranges of the data can be different. A good example is minutes and blocks per game - Most players will have more minutes per game than blocks per game.

```{r}
# scale/transform data
methods <- c('scale', 'log', 'cube_root')
# choose method - change index
method <- methods[1]
nba_feat_sc <- if (method == methods[1]) {
               scale(nba_feat) 
              } else if (method == methods[2]) {
              # replace 0s with small number
              nba_feat[nba_feat == 0] <- .0001
              log(nba_feat) 
              } else {
              (nba_feat)^(1/3)
              }

# sample size when sampling players in each cluster
# when taking the log, there may not be enough players to sample from
sample_size <- if (method == methods[2]) 2 else 10

# run PCA
# princomp() uses spectral decomposition
# prcomp() uses singular value decomposition
nba_pca <- prcomp(nba_feat_sc)
summary(nba_pca)
```

```{r}
# create plot PCA data function
plot_pca <- function(object, frame = FALSE, x = 1, y = 2, 
                     data, colour, title, label) {
  # plots data in PCA space 
  # object = PCA or K-Means object
  # x = which PC for x-axis (1, 2, ,3, etc..)
  # y = which PC for y-axis (1, 2, 3, etc..)
  # object: PCA or K-means object
  # data = underlying data
  p <- autoplot(nba_pca, x = x, y = y, data = nba, colour = colour, frame = frame) + 
        ggtitle(title) + 
        # center title
        theme(plot.title = element_text(hjust = 0.5)) + 
        geom_label_repel(aes(label = label),
                        box.padding   = 0.35, 
                        point.padding = 0.5,
                        segment.color = 'grey50') +
        theme_classic()
  return(p)
}
```


Since 2 components make up 83% of the cumulative variance, we will plot these

```{r}
# Labels: Players who played more than 36 min per game or less than 3 min per game
labels_pca <- ifelse(nba$MP_pg >= 36 | nba$MP_pg <= 3, 
                    as.character(nba$Player), '')
title_pca <- paste0('PCA: NBA - ', ncol(nba_feat) ,' features')

# Plot first two components with positions
plot_pca(nba_pca, data = nba, colour = 'Pos', 
         label = labels_pca, title = title_pca
          )
```


_Observations_
Based on the PCA plot, it looks like there are natural grouping based on position from the colors coding. For example, the centers are on the bottom diagonal, point guards are near the top, and SF/PFs are in the middle. It is also noticable that the stars and superstars are on the right-side of the cluster. Since this looks like a fan, we can also say that the stars are placed more towards the 'tips' of the fan. So, we will hypothesize that there may also be clusters from left to right, where the right-most are the top players, and the left side are the lower performing players.

## Clustering

### Hierarchical clustering
The first method of clustering we will try is hierarchical clustering. The dendrogram can help provide a visual aid in the number of clusters we can start to use.
```{r}
# distance matrix for features
nba_dist_sc <- dist(nba_feat_sc, method = 'euclidean')
  
# try single, centroid, and ward (D2) linkage hier clustering
hcl_single <- hclust(d = nba_dist_sc, method = 'single')
hcl_centroid <- hclust(d = nba_dist_sc, method = 'centroid')
hcl_ward <-  hclust(d = nba_dist_sc, method = 'ward.D2')
```

```{r}
# nearest neighbors method
plot(hcl_single, hang = -1, main = 'Single Linkage', labels = FALSE)
# groups centroid
plot(hcl_centroid, hang = -1, main = 'Centroid Linkage', labels = FALSE)
# Ward’s minimum variance method, 
# with dissimilarities are squared before clustering
dend <- as.dendrogram(hcl_ward)
hcl_k <- 4
dend_col <- color_branches(dend, k = hcl_k)
plot(dend_col, main = paste0('Ward (D2) Linkage: K = ', hcl_k))
```

Since the Ward dendrogram seems to be the best among the three, we will look at its distribution for 3 and 4 clusters. We chose these initial groupings because this is a good number of initial clusters to group NBA players.
```{r}
# add cluster labels to main data
nba$hcl_ward_labs_three <- cutree(hcl_ward, k = 3)
nba$hcl_ward_labs_four <- cutree(hcl_ward, k = 4)

# plot frequencies
p_one <- ggplot(data = nba, aes(x = hcl_ward_labs_three)) + 
          geom_bar(fill = 'lightblue') + ggtitle('HCL: K = 3')
p_two <- ggplot(data = nba, aes(x = hcl_ward_labs_four)) + 
          geom_bar(fill = 'lightblue') + ggtitle('HCL: K = 4')

# combine
cowplot::plot_grid(p_one, p_two, nrow = 2)
```


```{r}
# add labels to data
nba$hcl_ward_three <- factor(cutree(hcl_ward, k = 3))
nba$hcl_ward_four <- factor(cutree(hcl_ward, k = 4))

# player names to include in plot
hcl_labels <- ifelse(nba$MP_pg >= 36 | nba$MP_pg <= 2.5 |
                     (nba$MP_pg >= 28.8 & nba$MP_pg <= 29) | 
                     (nba$MP_pg >= 25 & nba$MP_pg <= 25.2),
                     as.character(nba$Player), '' )

# elements to loop over
hcl_labs <- names(nba %>% select(tail(names(.), 2)))
hcl_ks <- c(3, 4)

# plot hclust labels superimposed over PCA
for (i in seq_along(hcl_labs)) {
  p <- plot_pca(nba_pca, frame = TRUE, 
                data = nba, colour = hcl_labs[i],
                title = paste0('PCA: ', hcl_ks[i], ' clusters (Hclust)'),
                label = hcl_labels
  )

  print(p)
}
```


Visually, it looks like the four cluster solution may be able to give us more actionable insights vs the 3-cluster method. The average stats by cluster shows pretty clear separation among the groups. Group 4 are the stars, followed by group 2, 3, and then 4. The main difference between groups 2 and 3 is that group 2 looks to contain more players who tend to have more rebound and blocks per game.


```{r}
# averages by cluster
nba_hclust_avg <- data.frame(nba
                            %>% select(hcl_ward_four, MP_pg, PTS_pg, TRB_pg,
                                       AST_pg, BLK_pg, STL_pg, VORP, PER, RPM)
                            %>% group_by(hcl_ward_four)
                            %>% summarise_all(list(mean))
                            )
nba_hclust_avg
```


```{r}
# sample players from each cluster
for (k in 1:hcl_ks[2]) {
  print(paste0('Cluster ', k))
  print(sample(subset(nba, hcl_ward_four == k)$Player, size = sample_size))
}
```

### Optimize number of clusters

Method: Calinski-Harabasz index
```{r}
# get optimal cluster sizes 
cluster_sizes_hcl <- NbClust(data = nba_feat_sc,
                          # it will likely be harder to interpret clusters
                          # past this amount
                          max.nc = 6,
                          method = 'ward.D2',
                          index = 'ch')

# plot C(G)
plot(names(cluster_sizes_hcl$All.index),
     cluster_sizes_hcl$All.index,
     main = 'Calinski-Harabasz index: HCL',
     type = 'l')
```


Among the different hierarchical clustering methods, the Ward method seems to be the best. The dendogram looks the most structured and the distribution of players in each cluster is more balanced. Hierarchical clustering could seem like a potential fit if we want the better players to be in a more 'select' group. Although the CH index indicates 2 clusters is optimal, we need to look at the practicality as well. 3 clusters may hav differences between groups of players. But, it is possible that NBA front offices will likely need more differentation when grouping player performance. Looking at the 4 cluster solutions and stats, the blue cluster tends to have more players who rebound, block more shots and tend to be more efficient (based on PER). So, these clusters seem to have decent separation from each other. We will now try K-Means clustering too see if that works better.

### K-Means

```{r}
# try K-means clustering
# try arbitrary number of clusters first
km_k <- 5
km_five <- kmeans(x = nba_feat_sc,
            centers = km_k,
            nstart = 100,
            algorithm = 'Hartigan-Wong')

```


```{r}
# plot k-means clusters in PC space
nba$km_labs_five <- factor(km_five$cluster)
# Labels: Players who played more than 36 min per game or less than 3 min per game
km_labels <- ifelse(nba$MP_pg >= 36 | nba$MP_pg <= 3, 
                         as.character(nba$Player), '' )

plot_pca(km_five, data = nba, frame = TRUE, colour = 'km_labs_five',
          title = paste0('PCA: ', km_k, ' clusters (K-means)'),
          label = km_labels)
```


```{r}
# get distribution of players in each cluster
ggplot(data = nba, 
       aes(x = km_labs_five)) + 
       geom_bar(fill = 'lightblue') + 
       ggtitle('HCL: K = 5')

```


It looks like the clusters are somewhat interpretable. Clusters to the right seem to indicate star players, while clusters to the left indicate lower performing players. However, the question becomes if 5 clusters is meaningful. Based on the averages for each cluster, there is not much difference between clusters 2 and 4. Additionally, it looks like there is room to better balance the number of players in each cluster and create more separation between clusters. We will now optimize the number of clusters utilizing the Calinski-Harabasz index.

```{r}
# averages by cluster
nba_km_five_avg <- data.frame(nba
                            %>% select(km_labs_five, MP_pg, PTS_pg, TRB_pg,
                                       AST_pg, BLK_pg, STL_pg, VORP, PER, RPM)
                            %>% group_by(km_labs_five)
                            %>% summarise_all(list(mean))
                            )
nba_km_five_avg
```



### Optimize number of clusters

Method: Calinski-Harabasz index
```{r}
# get optimal cluster sizes 
cluster_sizes_km <- NbClust(data = nba_feat_sc,
                          # it will likely be harder to interpret clusters
                          # past this amount
                          max.nc = 6,
                          method = 'kmeans',
                          index = 'ch')

# plot C(G)
plot(names(cluster_sizes_km$All.index),
     cluster_sizes_km$All.index,
     main = 'Calinski-Harabasz index: K-Means',
     type = 'l')

```

```{r}
# get best number of clusters
cluster_sizes_km$Best.nc
```

```{r}
# show all indices
cluster_sizes_km$All.index
```

_Observations_
Although the CH index indicates that the optimal number of clusters is 2, this seems too low of a number to meaningfully break out the NBA players into groups. It is also important to note that the CH index is a heuristic method. So although CH is a good approach to look for the number of clusters, it is important to combine this with our practical goal of looking for underlying patterns in the players. Thus, I think a more reasonable number to understand the data is with 3 - 4 clusters, which show the second and third best partitions based on the CH index. We will look at both and determine which one is a better fit for our goal.


### Try 3 and 4 clusters (K-Means)

```{r}
# number of clusters to try
num_clust <- c(3, 4)

# store cluster results
clust_list <- list(rep(NA, length = length(num_clust)))
for (i in seq_along(num_clust)) {
  clust_list[[i]] <- kmeans(x = nba_feat_sc,
                      centers = num_clust[i],
                      nstart = 100,
                      algorithm = 'Hartigan-Wong')
            
  }
```


```{r}
# add cluster labels to main data
nba$km_labs_three <- factor(clust_list[[1]]$cluster)
nba$km_labs_four <- factor(clust_list[[2]]$cluster)
# elements to loop over
km_labs <- names(nba %>% select(tail(names(.), 2)))
km_ks <- c(3, 4)

# plot k-means clusters in PC space
for (i in seq_along(clust_list)) {
  p <- plot_pca(clust_list[[i]], data = nba, frame = TRUE, colour = km_labs[i],
          title = paste0('PCA: ', km_ks[i], ' clusters (K-means)'),
          label = km_labels)
  print(p)
}
```


```{r}
# number of players in each cluster
p_three <- ggplot(data = nba, aes(x = km_labs_three)) + 
          geom_bar(fill = 'lightblue') + ggtitle('KM: K = 3')
p_four <- ggplot(data = nba, aes(x = km_labs_four)) + 
          geom_bar(fill = 'lightblue') + ggtitle('KM: K = 4')

# combine
cowplot::plot_grid(p_three, p_four, nrow = 2)
```

We may be able to get more insight with four clusters instead of three. We will sample players from each cluster and look at cluster averages.
```{r}
# averages by cluster
nba_clust_avg <- data.frame(nba
                            %>% select(km_labs_four, MP_pg, PTS_pg, TRB_pg,
                                       AST_pg, BLK_pg, STL_pg)
                            %>% group_by(km_labs_four)
                            %>% summarise_all(list(mean))
                            )
nba_clust_avg
```

```{r}
# sample players in each cluster
for (k in 1:4) {
  print(paste0('Clusters: ', k))
  print(sample(subset(nba, km_labs_four == k)$Player, size = sample_size))
}

```


Based on the plot, cluster distributions, and group averages, it looks like 4 clusters is optimal. One main reason is that there is more separation vs 4 clusters, which can provide more value when bucketing players by overall skillsets. Across the statistics, it looks like the clusters are broken out into the following: Best players (2), Good players with more assists, i.e. guards (1), good players who rebound and block more, .e.g forwards (4), and Low-performing players (3). We will now try model-based clustering as a third method.

### Model-Based Clustering

```{r}
# run model-based clustering
# nba_mcl <- Mclust(nba_feat_sc)
# summary(nba_mcl)
```

## PITFALL: more minutes per game tends towards higher stats per game.. redo metric to per 36 min instead? However, this has its fallbacks too.. Ex: player who played for 2 min with 2 pts and 2 reb would transalte to 18 pts and 18 reb per 36 min.. ##

## Post-Cluster Analysis

We will now look at different statistics and demographics to see how the clustering lines up

### Clusters vs. Player Salaries
Try: K-means, 4 clusters
```{r}
# salary vs. advanced stats, overlayed with clusters

# cluster label to use
cl_label <- "km_labs_four"

# plot PER vs. salary
ggplot(data = nba, aes(x = Salary, y = PER)) + 
  geom_point(aes_string(color = cl_label)) + 
  geom_label_repel(aes(label = labels_pca),
                box.padding   = 0.35, 
                point.padding = 0.5,
                segment.color = 'grey50') + 
  ggtitle('PER vs. Salary') + 
  theme_classic()
```

```{r}
# plot salary distribution
ggplot(nba, aes(x = Salary)) + 
  geom_density(fill = 'lightblue', alpha = 0.4) + 
  ggtitle('Salary distribution') + 
  geom_vline(xintercept = mean(nba$Salary), 
             color = "blue", 
             linetype = 'dashed')

```

```{r}
# salary statistics
summary(nba$Salary)
```


```{r}
# Highest Paid players in Lowest Tier
data.frame(nba
           %>% select(Player, G, MP_pg, Tm, 
                      Salary, PER, cl_label)
           %>% filter(km_labs_four == 1)
           %>% arrange(desc(Salary))
           )

```

```{r}
# Lowest Paid players in highest tier
data.frame(nba
           %>% select(Player, G, MP_pg, Tm, Salary, PER, cl_label)
           %>% filter(km_labs_four == 4)
           %>% arrange(Salary)
           )

```


_Observations_
Based on the plot and tables, it looks like there is potential to update salaries based on player tiers. For example, Chandler Parsons was paid 22M but is considered a low tier player, and is paid more than the high tier players such as Steph Curry (12M) and Kawhi Leonard (17.6M).

### Team Compensation and Performance vs clusters

```{r}
# convert labels to numeric
nba$cl_lab_numeric <- as.numeric(nba[ , cl_label])
nba_team <- data.frame(nba 
                       %>% select(Tm, Salary, win_pct, cl_lab_numeric) 
                       %>% group_by(Tm) 
                       %>% summarise(team_salary = sum(Salary),
                                     win_pct = mean(win_pct),
                                     avg_clust = mean(cl_lab_numeric))
                       )

# order by descnding average cluster label
arrange(nba_team, desc(avg_clust))
```


```{r}
# plot
ggplot(nba_team, 
       aes(x = team_salary, y = win_pct, color = avg_clust)) + 
  geom_point() + 
  geom_label_repel(label = nba_team$Tm) + 
  ggtitle('Win % vs. Team Salary') + 
  theme_classic()
```


```{r}
# Inspect some teams
teams_sample <- c('NYK', 'GSW', 'BOS')

teams_sample_list <- list(rep(NA, length = length(teams_sample)))
for (i in seq_along(teams_sample)) {
  teams_sample_list[[i]] <- nba[nba$Tm == teams_sample[i], 
                                c('Player', 'PER', cl_label)]
}

# change index to see different teams
teams_sample_list[[2]]
```


_Observations_
Although a team can have better players on average clusters, there are many variables at play here. A team can be better on average but poor management or coaching can affect a team's overall performance, e.g. NYK. Interestingly, GSW did not have the highest average cluster rating, because their bench is not very strong. This speaks to the strong influence that starter players can have on team performance. Another interesting note is that teams can play well even if they do not have many all-stars or a strong overall team, e.g BOS. This could be driven by great coaching and team chemistry. It is important to note that items such as injuries could greatly influence win %, even if players have high ratings. 

Although there is a correlation between overall team salary and win %, it is interesting that average player rating does not necessarily align with overall win %.
